Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
no hyphen in hidden Markov

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

@@

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence of amino acids.
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
Mathematically, we represent a query sequence as
$\left\{x_{1}...x_{n}\right\}$.
Ideally, if we were writing something like the BioHaskell library, we would
have begun with an algebraic datatype for amino acids deriving
\texttt{Ix}, something like
\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
which would still allow us to index into an array or vector.
However, since we wanted to maintain compatibility with the file format
shared with previous software, we broke with idiomatic Haskell and represented
our amino acids as 
\begin{verbatim}
type AA = Int
\end{verbatim}
and thus our query sequence as
\begin{verbatim}
type querySeq = Vector AA
\end{verbatim}

The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{smallverbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{smallverbatim}
to represent the seven allowed state transitions.

>>>>

@@

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence of amino acids.
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
Mathematically, we represent a query sequence as
$\left\{x_{1}...x_{n}\right\}$.
Ideally, if we were writing something like the BioHaskell library, we would
have begun with an algebraic datatype for amino acids deriving
\texttt{Ix}, something like
\begin{verbatim}data Amino = Ala | Cys | Asp | Glu | Gly ...\end{verbatim}
which would still allow us to index into an array or vector.
However, since we wanted to maintain compatibility with the file format
shared with previous software, we broke with idiomatic Haskell and represented
our amino acids as 
\begin{verbatim}
type AA = Int
\end{verbatim}
and thus our query sequence as
\begin{verbatim}
type querySeq = Vector AA
\end{verbatim}

The states of the hidden Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state determines the probability that the query sequence
has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model, so it is a \textit{transition probability}.
Mathematically, transition probabilities are represented by $a_{s_{j-1}s_{j}}$
for a state $s_{j}$.
In Haskell, we use 
\begin{smallverbatim}
type logProb = Double
data TProb  = TProb { logProbability::LogProb
                    , fromState::HMMState
                    , toState::HMMState
                    }
\end{smallverbatim}
to represent the seven allowed state transitions.
